Type-Safe Modular Parsing

### 2017-01-09 paper/sections/Inheritance.tex
Changes
Changes on section 3, 4

@@

This section introduces the problem that motivates our work. In Figure~\ref{fig:packrat-arith} we have observed
that Packrat parsers can be defined in a convenient way, with recursive calls. The example is actually inspired
by Walder's Expression Problem. Similarly, our goal is to build parsers that are extensible in some dimensions.

\subsection{Dimensions of Extensibility}\label{subsec:dimensions}
We continue with the expression language of literals and additions. At this point,
one would like to have two kinds of extensions:

%\section{An Overview of \name}\label{sec:overview}
%
%This section gives an overview of our library \name, motivates
%additional problems that occur when parsing modular ASTs.  The first
%problem is that traditional parsers using parser combinators use
%hard-coded recursive calls. This problem is solved using
%delegation-based techniques, which enable parsing OO ASTs.  The second
%problem is how to achieve dual extensibility of ASTs (that is: how to
%allow for both new language constructs and new operations over the
%ASTs). This problem is solved using Object Algebras.

\begin{comment}
\subsection{Choosing the Parsing Technique}\label{subsec:overview-parsing}

%A technique for type-safe modular parsing should the following 3
%features: \emph{modular type-checking}; \emph{separate compilation};
%low performance overhead

In the last section, we have argued that parser combinators are a suitable parsing technique for
our purpose, as they naturally build modular parsers for type-checking.
Unfortunately many parser combinators have important limitations.
In particular several parser combinators,
including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination} and \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of them are
problematic in a modular setting.

\paragraph{Packrat Parsing}
Fortunately some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated.
Moreover, theoretically the algorithm behind Packrat parsers
supports both direct and indirect left-recursion~\cite{warth2008}.
The current version of the library is still
buggy with indirect left-recursion, but we believe that
they will fix it in the future, and direct left-recursion is
already practical to use for a large number of applications. All of these properties are very
suitable for modularity, thus we decided to use Packrat parsers as the underlying
parsing technique in \name.

It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}

%\huang{done, added the attempt/failure/reason of extending conventional parsers}\bruno{You are not motivating the problem! You are going straight to
%  the solution without pointing out what the problem is first. What
%  you need to do is: First show what happens with conventional
%  parsers: at some point, if you add extensions the recursive calls
%  will be wrong. Then you show (in the next section) the solution:
% use delegation/open recursion.}

%\huang{partly done, only extend a new language construct in the example, how about leaving extension of new operations in the OA subsection?}\bruno{I think we need to set up a challenge here, similar to the challenge of
%  the expression problem. The challenge should be like. Build a parser
%for a simple expression language, then extend the language with both
%a new language construct and a new operation. This section will show
%how we can do that using traditional parsers, but without modular
%type-safety and separate compilation. The remaining sections will show
%that the techniques introduced by us, enable us to solve those two challenges.}

\begin{itemize}
\item \textbf{Extension 1}: adding a new case (or rather, a new construct) to \inlinecode{Expr};
\item \textbf{Extension 2}: adding a new operation to \inlinecode{Expr}.
\end{itemize}

\paragraph{Attempt to Extend the Parser} Now we first consider
extension 1, namely we are extending the syntax as well as the parser. Hence we introduce variables as a new case.

@@

As in an OO solution to the Expression Problem, it is easy to extend
the corresponding AST:

@@

And since we already have the parser for literals and additions, we would
like to build the new parser by reusing the old one. Here one may quickly come up
with the following attempt, where a new parser is defined for \inlinecode{Var}, then composed
with \inlinecode{pExpr}:

\huang{Extending from old parser is necessory to get the `pExpr', added a surrounding trait.}
\lstinputlisting[linerange=35-38]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_BAD_ATTEMPT

Unfortunately, it fails to parse some expressions like \inlinecode{"1 + x"}, which is obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. Still, one may continue to change the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr},
but modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

\subsection{Inheritance-Based Approach}\label{subsec:inheritance-approach}

\huang{It seems a little bit weird since we were already using inheritance in the `bad attempt'. I think we could focus on overriding here}

Fortunately in Scala we can make use of inheritance for modular parsing. Specifically,
the new parser can be defined in an enclosing trait that extends
the old one:

\lstinputlisting[linerange=42-47]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH
Now \inlinecode{VarExprParser} successfully represents the parser for the extended language, because Scala uses dynamic dispatch for
method overriding in inheritance. When the input \inlinecode{"1 + x"} is fed to the parser \inlinecode{this.pExpr}, it firstly delegates
the work to \inlinecode{super.pExpr}, which parses literals and additions. However, the recursive call \inlinecode{pExpr} in \inlinecode{pAdd}
actually refers to \inlinecode{this.pExpr} again, due to dynamic dispatch. It implies that the recursive call has included the extension. This
elegant inheritance-based approach opens the recursion of parsers in a modular way.

@@

\lstinputlisting[linerange=56-62]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

As we demonstrated, inheritance is the key technique to obtain semantic modularity.
It enables type-safe code reuse and separate compilation for parsing OO style ASTs with extension 1.

>>>>

\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}

@@

This section introduces the problem of semantic modular parsing which motivates our work, and an initial solution using only standard inheritance in OO.

\subsection{Modular Parsing Problem}\label{subsec:parsingproblem}
\huang{or "Parsing Extensible ASTs" ?}
The extensibility issues of AST structures and operations that process them, can be illustrated by the famous Expression Problem~\cite{wadler1998expression}. There are two dimensions of extensibility:

\begin{itemize}
\item \textbf{Extension 1}: adding a new data variant (or rather, a new constructor of ASTs).
\item \textbf{Extension 2}: adding a new operation over ASTs.
\end{itemize}

We already have solutions such Object Algebras~\cite{Oliveira2012} for this problem. However, when ASTs evolve, the corresponding parsers which produce such structures should also change accordingly. Our motivation is to build modular parsers, which can be extended together with ASTs. Futhermore, we focus on semantic modularity, that means we expect parsers to be modularly type-checked and separately compiled.

In the rest of Section~\ref{sec:inheritance}, we will introduce an solution for modular parsing with Extension 1 above, using only standard inheritance in OO. Extension 2 will be discussed in Section~\ref{sec:algebrasandparsing}.

\subsection{Inheritance-Based Parser}

Let's firstly consider Extension 1, which is to extend the syntax as well as the parser. As an example, we continue with the expression language of literals and additions in Section~\ref{subsec:packratparsing}. We introduce variables as a new case as below.

@@

It is easy to extend the corresponding AST in an OO way:

@@

Since we already have the parser for literals and additions, we would
like to build the new parser by reusing the old one. In Scala we can take advantage of inheritance for such reuse. Specifically,
the new parser can be defined in an enclosing trait that extends
the old one. Here one may quickly come up
with the following attempt, where a new parser is defined for \inlinecode{Var}, then composed
with \inlinecode{pExpr}:

\lstinputlisting[linerange=35-38]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_BAD_ATTEMPT

Unfortunately, it fails to parse some expressions like \inlinecode{"1 + x"}, which is obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

\subsection{Overriding for Extensibility}

It is actually quite simple to let \inlinecode{pExpr} cover the newly extended case without modifying existing code. Method overriding is a standard feature which often comes with inheritance, and it allows us to explain inherited method again, such as \inlinecode{pExpr}. We can build the new parser which correctly parses \inlinecode{"1 + x"} as below.

\lstinputlisting[linerange=42-47]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH

Now \inlinecode{VarExprParser} successfully represents the parser for the extended language, because Scala uses dynamic dispatch for
method overriding in inheritance. When the input \inlinecode{"1 + x"} is fed to the parser \inlinecode{this.pExpr}, it firstly delegates
the work to \inlinecode{super.pExpr}, which parses literals and additions. However, the recursive call \inlinecode{pExpr} in \inlinecode{pAdd}
actually refers to \inlinecode{this.pExpr} again due to dynamic dispatch, and it covers the variable case. Similarly, all recursive calls can be updated to include new extensions if needed.

@@

\lstinputlisting[linerange=70-76]{../Scala/Parser/src/PaperCode/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

As we demonstrated, inheritance is the key technique to obtain semantic modularity.
It enables type-safe code reuse and separate compilation for parsing OO style ASTs with Extension 1.

### 2017-01-09 paper/sections/ObjectAlgebras.tex
Changes on section 3, 4

The inheritance-based approach benefits us from building extensible parsers, based on an OO class hierarchy.
Nevertheless, the addition of new operations over the ASTs becomes problematic. Often it requires modification on
the existing code, which breaks modularity. In this section, we will have a closer look at how it motivates us to
use Object Algebras for full extensibility.

@@

Now it comes to extension 2 in Section~\ref{subsec:dimensions}. In the previous code, we have presented pretty-printing as
an example of operations on the ASTs. Suppose now we also want to collect free variables on the expression language, which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

\lstinputlisting[linerange=19-19]{../Scala/Parser/src/PaperCode/Sec4OA/Code1.scala}% APPLY:linerange=ATTEMPT_EXPRWITHFREEVARS
It refers to one Scala solution to the Expression Problem []. But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \inlinecode{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

In contrast, Object Algebras~\cite{Oliveira2012} enables us to solve this problem. As a nice solution to the Expression Problem, it separates data variants
and operations, and offers high flexibility in the choice of operations to be performed over the AST. Before adopting it for parsing, the next section introduces the background of Object Algebras.

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem nicely,
achieving two dimensions of extensibility (data variants and operations) in a modular and type-safe way.
Because of this, the definition of data structures is separated from behaviors on them, and future extensions
to both sides no longer require existing code to be modified, supporting separate compilation.

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The example from Section~\ref{subsec:packratparsing} is again used here for illustration.
At first the language only supports literals and additions:

\lstinputlisting[linerange=5-8]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
The language has two constructors: literals and additions. The trait \inlinecode{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \inlinecode{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

\lstinputlisting[linerange=12-15]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
Furthermore, one can define a new trait to implement evaluation on arithmetic expressions:

@@

On the other hand, the data variants can be extended by inheriting \inlinecode{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:\haoyuan{Can we use var instead of varE?}

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
We observe that only the new case is implemented for pretty-printing, and the others have been inherited.
It offers programmers with great code reuse, in the sense that existing cases can also be
overridden in \inlinecode{VarExprPrint} using \inlinecode{"override def"}.

On this, to create an expression of \inlinecode{VarExprAlg}, a generic method is defined as follows:

>>>>

The inheritance-based approach benefits us from building extensible parsers, based on an OO class hierarchy.
Nevertheless, the addition of new operations over ASTs is problematic using traditional OO classes. Often it requires modification on existing code, which breaks modularity. In this section, we will have a closer look at how Object Algebras is applied for full extensibility.

@@

Now it comes to Extension 2 in Section~\ref{subsec:parsingproblem}. In the previous code, we have presented pretty-printing as
an example of operations on ASTs. Suppose we also want to collect free variables on the expression language which models
literals, additions and variables. Based on the traditional OO class hierarchy, one attempt would be extending \inlinecode{Expr} with the new
operation directly:

\lstinputlisting[linerange=19-19]{../Scala/Parser/src/PaperCode/Sec4OA/Code1.scala}% APPLY:linerange=ATTEMPT_EXPRWITHFREEVARS

It refers to one Scala solution to the Expression Problem.\huang{what's the solution? is it successful?} But then programmers have to define new classes for literals, additions and variables
by extending the old ones together with \inlinecode{FreeVarsExpr}. The critical issue is that, the existing parsing code only generates objects of the old
types, again it is a modularity issue, since it is a bad idea to modify existing code again and again with extensions. Other solutions including modelling constructors as functions introduce unnecessary complexity.

In contrast, Object Algebras~\cite{Oliveira2012} enables us to solve this problem. As a nice solution, it separates data variants
and operations, and offers high flexibility in the choice of operations to be performed over ASTs. Before adopting it for parsing, the next section introduces the background of Object Algebras.

\subsection{Object Algebras}\label{subsec:objectalgebras}
\huang{to polish}Object Algebras captures a design pattern to address the Expression Problem nicely,
achieving two dimensions of extensibility (data variants and operations) in a modular and type-safe way.
Because of this, the definition of data structures is separated from behaviors on them, and future extensions
to both sides no longer require existing code to be modified, supporting separate compilation.

Using Object Algebras in Scala, ASTs as recursive data structures are defined by traits, where each constructor corresponds to an abstract method inside. The example from Section~\ref{subsec:packratparsing} is used here again for illustration.
At first the language only supports literals and additions:

\lstinputlisting[linerange=5-8]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \inlinecode{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

\lstinputlisting[linerange=12-15]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression langauge:

@@

Futhermore, the data variants can be extended by inheriting \inlinecode{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:

@@

\lstinputlisting[linerange=32-34]{../Scala/Parser/src/PaperCode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
We observe that only the new case is implemented for pretty-printing, and the others have been inherited.
It offers programmers great code reuse, in the sense that existing cases can also be
overridden in \inlinecode{VarExprPrint}.

To create an expression of \inlinecode{VarExprAlg}, a generic method is defined as follows:
